### 2022-02-18 paper/paper.tex
adding paper, finishing model in rough state

\documentclass{article}
\usepackage{amsmath}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue,
    pdftitle={Overleaf Example},
    pdfpagemode=FullScreen,
}

\begin{document}

\title{Do professional chess players draw too often?}
\author{Tudor Muntianu}
\maketitle

\section*{Disclaimer}
I spent most of my time trying to get the models working. It was a complete nightmare. Getting the factor analysis model to work (replacing Smerdon's PCA) took me about 10 hours.
Maybe I should have settled for PCA for this draft.
Perhaps I should not have tried to relearn Pyro, a probabilistic programming language written in Python.
That being said, if I had to build this model in R I likely would have ended up using RStan, which is a binding to another PPL, so the model wouldn't have really been in R either.
The data cleaning and manipulation would also have taken me 10x longer since I don't really know R.
Hindsight is 20/20.

The code (and this writeup) is in a rough state as a result -- by the final draft it will be pretty and commented and clear (despite being in Python).
Similarly, since the model was giving me issues I didn't have compute time to train the model on the whole dataset (which is 4.6M games).
Again this will not be the case by the final draft. Everything works though; I have a complete pipeline
that just needs some clean up, some hyperparameters to be adjusted/played with, and to be run to completion.

Also, there are a couple of things I would like to add (or at least investigate) by the final draft:
\begin{itemize}
    \item Dataset of online games (ostensibly less incentive to draw here, more ``naturalistic'' or aggressive gameplay)
    \item Using the factor model to compute an index, comparing to Smerdon's results
\end{itemize}

All code can be found \href{https://github.com/tmuntianu/chessdraws}{here}.

\section{Introduction}

The prevalence of draws in professional chess has been steadily increasing over the past couple decades.
As play becomes more precise and accurate due to extensive computer preparation and analysis,
and as more money enters the sport, winning edges are diminished and players are increasingly
incentivized not to lose.
For instance, at the 2018 World Chess Championship, where Magnus Carlsen played Fabiano Caruana (World No. 1 and No.2 respectively),
all 12 classical time-control games were drawn.
The championship was decided by a rapid chess tiebreaker.
Near-perfect play is not the only factor, however.
Since tournament fates, prize pools, and sponsorship deals hinge on performance, losing is highly undesirable.
So, the popular argument goes, players play for and agree upon too many draws. This opinion is shared by
many strong players and professionals, notably Susan Polgar, a very strong grandmaster who argues that agreed draws should be eliminated from the game.
While some agreed draws are legitimate reflections of drawn game state, others occur when players simply play out opening moves and subsequently agree to draw
to minimize variance. This sort of behavior is most egregious, but it demonstrates the most extreme form of this behavior, making it likely
that playing passively or ``soft-playing'' for draws also occurs.

The question we want to answer is: are draws in high-level chess a result of optimal play?
or are players sacrificing chances to win to reduce variance as a result of external pressures?
Playing ``solid'' chess (i.e., slow, methodical, less aggressive) that lends itself to more draws is not what we are investigating,
but rather if players are playing worse as a result of draw incentives. In other words, are players
playing below their true strengths as a result of playing for draw?
Put even another way, are players are playing to maximize the single-game outcome (i.e., win if possible, draw if losing),
or are players sacrificing performance to reduce their variance?
% From here out, irrational play will be used to refer to this playstyle. This is not technically true, since the payoffs for players are not just winning or losing, but reputation, sponsorships, tournament placement, etc.
% But for the purposes of this paper, playing rationally means playing to maximize the outcome of an individual game, not playing to maximize the real-world consequences.

In this paper the frequency of draws is estimated and compared to existing estimates, namely the very robust Elo rating system used in professional play. We begin by adapting David Smerdon's Fighting Chess Index to generate latent variables representing player behavior and aggression,
then use those latent variables in a simple Bayesian model to estimate the probability of winning, drawing, and losing.
Then, using the Elo rating system, we compute the expected value of the game's outcome, and compare our estimated probabilities to those generated by ratings to determine whether drawing behavior results in poorer results.

\section{The game}

In this paper we make a number of assumptions about each chess game observed.
First, we assume that each player plays at the strength of their respective Elo rating, or
in other words, plays their best. We make no attempt to model player strength directly or
measure the variance in their performance.
Instead, we want to know how aggressive or how passive they are playing.

Put more formally, given two players $W$ and $B$, we want to know
$p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$, where $p_{\text{outcome}}$ is
the probability of $W$ (white) winning, drawing, and losing respectively.

The Elo system, while it does not give a perfect estimate of $p$, does shed some light.

\subsection{Elo ratings}
The Elo rating system gives an excellent estimate of the strength of a player.
With sufficient games played against many opponents, its application to chess produces robust results. All major chess governing bodies (USCF and FIDE) use Elo as their rating system,
chess engines are benchmarked using Elo, and it is used ubiquitously in tournaments to determine seeding and matches.

Given the ratings $R_A$ and $R_B$ of two players $A$ and $B$ respectively, the Elo rating system functions as follows.
\begin{align*}
    E_A = \frac{1}{1+10^{(R_B-R_A)/400}} \\
    \text{where } E_A = p_{\text{win}} + \frac{1}{2} p_{\text{draw}}
\end{align*}
It follows from this that $E_A+E_B=1$, unsurprisingly,

\subsection{Player strategies and equilibrium}

The Elo rating system makes an implicit assumption that draw frequency does not directly affect
the expected score $E$.
There are infinite solutions to $E = p_{\text{win}} + \frac{1}{2} p_{\text{draw}}$.
So, to simplify chess for our purposes, we assume that each player plays
according to their rating, and the choice they must make is how aggressive or passive to play.

If we assume, for instance, that both players $W$ and $B$, are playing at maximal aggression, then $E_W=p_{\text{win}}$ and $E_B=p_{\text{lose}}$. That is, $p_{\text{draw}}=0$.
But if both players are not maximally aggressive, the probability of a draw increases.
To write out all the strategies, we could define an index $a_W, a_B$ for each player that indicates their aggression,
and then rebalance
$p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$ accordingly, but enumerating the strategies formally turns out not to matter.
This is because, under Elo assumptions, $E_W$ and $E_B$ are constant, regardless of draw probability.
So while players can change the probability of drawing depending on their strategies, they
cannot change their expected score.
% below is not correct, fix the math
% Both players can contribute to this probability, so $p_{\text{draw}}=p_{\text{draw}_W}+p_{\text{draw}_B}$,
% where $p_{\text{draw}_W}$ represents $W$'s passive play or propensity to draw.
% Under Elo assumptions, without changing their expected score, a player $A$ can choose to reduce
% $p_{\text{win}_A}$ by $f$ to increase $p_{\text{draw}_A}$ by $2f$ (and therefore increase $p_{\text{draw}}$ as well).
%
% Because this adjustment can result in $p_{\text{draw}}>1$, we limit each player's contribution to $0.5$, that is
% $p_{\text{draw}_A}<0.5$.
% This allows the total probability of a draw $p_{\text{draw}}$ to be $1$, which while unrealistic

This makes any set of strategies chosen by both players a Nash equilibrium under Elo assumptions, since
the expected score for any two strategies is equal.
Intuitively, under Elo's system, players can play well
aggressively and they can play well passively; the choice of playstyle will not affect the outcome.

Our models will estimate these probabilities $p$ in order to determine whether these draw probabilities are in accordance with Elo's system, that is, reflective of player ability and playstyle,
or if they result from ``irrational play.''

The prediction, as described by Polgar \href{https://chessdailynews.com/the-700-lbs-gorilla-issue-to-draw-or-not-to-draw/}{here}, is that real-world circumstances encourage drawing at the expense of good chess.
As a result of chess being played sub-optimally, we would expect that as a result of a player playing for a draw more aggressively,
they sacrifice some results. In other words, if you play to minimize variance, you decrease your expected score.

\section{The models}
The dataset used was Caissabase, a free database of over 5M professional chess games.
Simul games, Chess960 games, and consultation (team) games were removed.
Games played by players with less than 10 games were also removed.

(I will make better citations and give more background and explanations of these models by the final draft I promise. I am rushing at the moment.)

\subsection{Player model}
For modelling player behavior, a factor model was used, inspired by David Smerdon's
\href{https://www.davidsmerdon.com/?p=2168}{Fighting Chess Index}.
Variables computed for each player were
\begin{itemize}
    \item total games
    \item average rating
    \item average absolute difference between player ratings
    \item average number of draws
    \item average number of draws in 30 moves or less
    \item average number of draws in 30 moves or less when player was White
    \item average number of moves in drawn games
\end{itemize}
See Section 12.2.4 in Bishop's \textit{Pattern Recognition and Machine Learning} for a detailed description of this factor analysis model.

Smerdon uses PCA to form his Fighting Chess Index, but this method has its flaws. As described in Bishop,
PCA can be formulated as a maximum likelihood solution of a latent variable model.
But PCA does not treat the observed variables as conditionally independent given the latent variables.
So if we use principal components to measure ``aggression'' or ``fighting spirit,'' we are not treating
average rating (a measure of player strength) and irrational draw behavior as independent.
This is problem since they are correlated.
More serious (stronger) players have more on the line, and so will generally play for draws more often.
But as we want to measure ``fighting spirit'' for individual player independent of their strength, we do not want this dependence.

In Smerdon's defense, PCA's components do lend themselves to the construction of his index (whereas the factor model used here does not), but this does not change the problem of the underlying assumption.

The factor model used here does treat observations as conditionally independent of the latent variables.
This is tremendously useful, since we want to retain measures of player strength while independently encoding
draw behavior, especially as inputs to the following model.

\subsection{Game model}
(Again I will provide more detail wrt inference, etc in final, in a rush to the deadline as a result of repeated disasters)

For each game, we modelled $p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$
using the following generative model, where $l_w, l_b$ are the latent variables
 computed from the player model for the corresponding players.
\begin{align*}
    p = A l_w + B l_b + \epsilon \\
    \epsilon \sim \mathcal{N}(\mathbf{0}, \sigma\mathbf{I}) \\
    \sigma\sim \Gamma(\alpha=1,\beta=2)
\end{align*}

\section{Results}

\end{document}

### 2022-02-18 paper/paper.tex
more of paper

@@
 For each game, we modelled $p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$
-using the following generative model, where $l_w, l_b$ are the latent variables
- computed from the player model for the corresponding players.
+using the following Bayesian generative model, where $l_w, l_b$ are the latent variables
+ computed from the player model for the corresponding players, and $r$ is the result of the game.
 \begin{align*}
-    p = A l_w + B l_b + \epsilon \\
-    \epsilon \sim \mathcal{N}(\mathbf{0}, \sigma\mathbf{I}) \\
-    \sigma\sim \Gamma(\alpha=1,\beta=2)
+    \sigma &\sim \Gamma(\alpha=1,\beta=2) \\
+    \epsilon  &\sim \mathcal{N}(\mathbf{0}, \sigma\mathbf{I}) \\
+    p  &= A l_w + B l_b + \epsilon \\
+    p_{\text{pos}}  &= \text{Softplus}(p)= \log(1+\exp(p)) \\
+    p_{\text{norm}}  &= \frac{p_{\text{pos}}}{\sum_{i=1}^3 p_{\text{pos}_i}} \\
+    r  &\sim \text{Cat}(3, p)
 \end{align*}
 
+The result $r$ is drawn from a categorical distribution parameterized by $p_{\text{norm}}$, which in turn is a normalized linear transformation of the latent variables $l_w, l_b$ representing the players.
+The parameters are optimized via stochastic variational inference using a family of multivariate Gaussians
+as the variational distribution.
+I will also try MAP estimation and see the effect on the parameters.
+
+Once matrices $A$ and $B$ are found, we can compute $p_{\text{norm}}$ for each game from the latent variables.
+This can then be transformed into $\hat{E}_W$ and $\hat{E}_B$.
+Similarly, we can compute $E_W, E_B$ from player Elo ratings, as described previously.
+
 \section{Results}
+As of now I did not have enough time to train the model on the full dataset (or even a significant subset).
+
+However, to interpret/compare the results, here a few ideas (a couple of which are quickly explored in the notebooks, but mostly unimplemented):
+\begin{itemize}
+    \item Compare $\sum_{i=1}^{ngames}E_{W_i}$,
+        $\sum_{i=1}^{ngames}\hat{E}_{W_i}$, and
+        $\sum_{i=1}^{ngames}r_i$ where $r_j$ is the result of game $j$, and $r=1$ if $W$ wins and $r=0.5$ if it was a draw.
+        This is not a very informative measure, but should express whether the expected scores over- or under-estimate the actual scores.
+    \item Rank players by average draw probability to form an analogous index, compare with Smerdon's index
+    \item Check to see if $\hat{E}_{W_i}$ predicts upsets(victories against players with significantly higher ratings), and if it does so correctly, investigate why
+    \item Visualize variance in latent variables as well as in the generated features
+    \item Compare/regress draw probabilities across different ratings and/or different differences in rating
+    \item Split dataset into training and testing, see if it correctly predicts draws on the test dataset
+\end{itemize}
+
+These (future) results can relate back to the article/real-world situation in a couple of ways.
+First, they will show whether or not high draw rates are as a result of sub-optimal play rather than higher accuracy at high levels of chess.
+This would indicate that other pressures induce players into actually playing worse chess to avoid variance.
+This would provide a new angle from which to attack the problem of increasing draws.
+
+The article mentions measures designed to increase errors and decrease the likelihood of draws.
+These would likely be successful at the highest echelons of play, where games are sufficiently
+scrutinized and play is so precise that suboptimal play essentially does not exist.
+However, at lower levels of play (still professional, but just among weaker players), the problem is not absolute precision.
+Rather, (assuming the results support this), the problem is this variance-reducing behavior.
+So from a design perspective, the solution would be to penalize this behavior.
+Making draws worth $\frac{1}{3}$ of a point could discourage taking draws in tournaments, for instance.
+Removing agreed draws (where players simply agree to a draw rather than playing it out) as a whole could also make this behavior more difficult.
 

### 2022-02-18 paper/paper.tex
final edits

@@
 \section*{Disclaimer}
-I spent most of my time trying to get the models working. It was a complete nightmare. Getting the factor analysis model to work (replacing Smerdon's PCA) took me about 10 hours.
+I spent most of my time trying to get the models working. It was a complete nightmare. I thought it was close to working for many hours. Last 1\% took 99\% of the time.
+Getting the factor analysis model to work (replacing Smerdon's PCA) took me about 10 hours alone.
 Maybe I should have settled for PCA for this draft.
@@
 The data cleaning and manipulation would also have taken me 10x longer since I don't really know R.
-Hindsight is 20/20.
+Hindsight is 20/20. I think it's on its way, though.
 
 The code (and this writeup) is in a rough state as a result -- by the final draft it will be pretty and commented and clear (despite being in Python).
-Similarly, since the model was giving me issues I didn't have compute time to train the model on the whole dataset (which is 4.6M games).
+Similarly, since the model was giving me issues I didn't have compute time to train the model on the whole dataset (which is 4.6M games after cleaning).
 Again this will not be the case by the final draft. Everything works though; I have a complete pipeline
@@
 scrutinized and play is so precise that suboptimal play essentially does not exist.
-However, at lower levels of play (still professional, but just among weaker players), the problem is not absolute precision.
+However, at slightly lower levels of play (still professional, but just among weaker players), the problem is not absolute precision.
 Rather, (assuming the results support this), the problem is this variance-reducing behavior.

### 2022-03-08 paper/paper.tex
adding rewrite

--- 2022-02-18 paper/paper.tex
@@
 \usepackage{hyperref}
+\usepackage{graphicx}
 \hypersetup{
@@
     urlcolor=blue,
-    pdftitle={Overleaf Example},
-    pdfpagemode=FullScreen,
 }
+\usepackage[parfill]{parskip}
+\usepackage{bm}
+
+\newcommand{\pw}{p_{\text{win}}}
+\newcommand{\pd}{p_{\text{draw}}}
+\newcommand{\pl}{p_{\text{lose}}}
+\newcommand{\Rw}{R_{W_{\text{adj}}}}
+\newcommand{\Rb}{R_{B_{\text{adj}}}}
 
@@
 
-\section{The game}
+\section{Optimal chess}
 
-In this paper we make a number of assumptions about each chess game observed.
-First, we assume that each player plays at the strength of their respective Elo rating, or
-in other words, plays their best. We make no attempt to model player strength directly or
-measure the variance in their performance.
-Instead, we want to know how aggressive or how passive they are playing.
-
-Put more formally, given two players $W$ and $B$, we want to know
+Given two players $W$ and $B$, we want to know
 $p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$, where $p_{\text{outcome}}$ is
@@
 
-Given the ratings $R_A$ and $R_B$ of two players $A$ and $B$ respectively, the Elo rating system functions as follows.
-\begin{align*}
-    E_A = \frac{1}{1+10^{(R_B-R_A)/400}} \\
+Given the ratings $R_A$ and $R_B$ of two players $A$ and $B$ respectively, the Elo rating system predicts game outcomes as follows.
+\begin{gather*}
+    Q_A = 10^{R_A/400}, Q_B = 10^{R_B/400} \\
+    E_A = \frac{Q_A}{Q_A+Q_B} \\
     \text{where } E_A = p_{\text{win}} + \frac{1}{2} p_{\text{draw}}
-\end{align*}
-It follows from this that $E_A+E_B=1$, unsurprisingly,
+\end{gather*}
+It follows from this that $E_A+E_B=1$.
 
-\subsection{Player strategies and equilibrium}
+\subsection{Equilibria under Elo assumptions}
 
 The Elo rating system makes an implicit assumption that draw frequency does not directly affect
-the expected score $E$.
+the expected score $E$, as it does not define $p_{\text{win}}$ directly.
+
 There are infinite solutions to $E = p_{\text{win}} + \frac{1}{2} p_{\text{draw}}$.
-So, to simplify chess for our purposes, we assume that each player plays
-according to their rating, and the choice they must make is how aggressive or passive to play.
+Under Elo assumptions players can play solid passive chess and solid aggressive chess,
+changing the probabilities of winning and drawing but importantly \textit{not changing}
+the total expected value $E$.
 
-If we assume, for instance, that both players $W$ and $B$, are playing at maximal aggression, then $E_W=p_{\text{win}}$ and $E_B=p_{\text{lose}}$. That is, $p_{\text{draw}}=0$.
-But if both players are not maximally aggressive, the probability of a draw increases.
-To write out all the strategies, we could define an index $a_W, a_B$ for each player that indicates their aggression,
-and then rebalance
-$p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$ accordingly, but enumerating the strategies formally turns out not to matter.
-This is because, under Elo assumptions, $E_W$ and $E_B$ are constant, regardless of draw probability.
-So while players can change the probability of drawing depending on their strategies, they
-cannot change their expected score.
-% below is not correct, fix the math
-% Both players can contribute to this probability, so $p_{\text{draw}}=p_{\text{draw}_W}+p_{\text{draw}_B}$,
-% where $p_{\text{draw}_W}$ represents $W$'s passive play or propensity to draw.
-% Under Elo assumptions, without changing their expected score, a player $A$ can choose to reduce
-% $p_{\text{win}_A}$ by $f$ to increase $p_{\text{draw}_A}$ by $2f$ (and therefore increase $p_{\text{draw}}$ as well).
-%
-% Because this adjustment can result in $p_{\text{draw}}>1$, we limit each player's contribution to $0.5$, that is
-% $p_{\text{draw}_A}<0.5$.
-% This allows the total probability of a draw $p_{\text{draw}}$ to be $1$, which while unrealistic
-
-This makes any set of strategies chosen by both players a Nash equilibrium under Elo assumptions, since
-the expected score for any two strategies is equal.
+This makes any set of strategies chosen by both players a Nash equilibrium under Elo assumptions, since the expected score for any two strategies is equal.
 Intuitively, under Elo's system, players can play well
@@
 
-Our models will estimate these probabilities $p$ in order to determine whether these draw probabilities are in accordance with Elo's system, that is, reflective of player ability and playstyle,
-or if they result from ``irrational play.''
+But if players play an overly passive style to achieve more draws,
+they decrease their total expected value $E$ to increase $p_{\text{draw}}$.
+This is the sort of behavior we are interested in, whether players play
+too passively or draw too often \textit{at the expense} of playing optimal chess.
 
-The prediction, as described by Polgar \href{https://chessdailynews.com/the-700-lbs-gorilla-issue-to-draw-or-not-to-draw/}{here}, is that real-world circumstances encourage drawing at the expense of good chess.
+This is Polgar's prediction, as described \href{https://chessdailynews.com/the-700-lbs-gorilla-issue-to-draw-or-not-to-draw/}{here}; i.e., real-world circumstances encourage drawing over playing optimally.
 As a result of chess being played sub-optimally, we would expect that as a result of a player playing for a draw more aggressively,
 they sacrifice some results. In other words, if you play to minimize variance, you decrease your expected score.
+To describe this, we introduce a modified Elo system.
 
-\section{The models}
-The dataset used was Caissabase, a free database of over 5M professional chess games.
-Simul games, Chess960 games, and consultation (team) games were removed.
-Games played by players with less than 10 games were also removed.
+\section{(Potentially) suboptimal chess}
 
-(I will make better citations and give more background and explanations of these models by the final draft I promise. I am rushing at the moment.)
+First we describe the framework for describe hyperpassive, suboptimal chess before discussing potential equilibria.
 
-\subsection{Player model}
-For modelling player behavior, a factor model was used, inspired by David Smerdon's
+\subsection{Modified Elo game}
+
+There are two players, $W, B$, white and black, with ratings $R_W,R_B$ respectively.
+
+Each chooses how ``overly passive'' to play, that is, how much they want to
+sacrifice expected value for increased draw percentage.
+
+For both $W,B$, their strategies are to choose $$\alpha_W, \alpha_B \in [0,1]$$
+
+Their payoffs, $\Pi_W,\Pi_B$, are represented by the following modified Elo equations.
+Similar to the standard Elo equations above, they represent the expected score of each player,
+but instead the expected score is scaled by $\alpha_W, \alpha_B$.
+% , and the probability of winning is scaled by a sigmoid-weighted average of $\alpha_W,\alpha_B$.
+\begin{gather*}
+    % \sigma(x)=\frac{1}{1+\exp(-x)} \\
+    Q_W = 10^{R_W/400}, Q_B = 10^{R_B/400} \\
+    \Pi_W = \frac{\alpha_W Q_W}{\alpha_W Q_A+\alpha_B Q_B}
+    % = \frac{\sigma(\alpha_W)\alpha_W + \sigma(\alpha_B)\alpha_B}
+    % {\sigma(\alpha_W) + \sigma(\alpha_B)}
+    % P(W\text{ Win}) + \frac{1}{2} P(\text{Draw})
+    \\
+    \Pi_B = \frac{\alpha_B Q_B}{\alpha_W Q_A+\alpha_B Q_B}
+    % = \frac{\sigma(\alpha_W)\alpha_W + \sigma(\alpha_B)\alpha_B}
+    % {\sigma(\alpha_W) + \sigma(\alpha_B)}
+    % P(B\text{ Win}) + \frac{1}{2} P(\text{Draw})
+\end{gather*}
+
+If a player $A$ chooses $\alpha=1$, then they play optimally, that is
+$\alpha_A Q_A = Q_A$.
+If they choose $\alpha=0$ then they ``turtle,''
+give up all chances of winning only to increase the likelihood of drawing.
+Importantly, if $\alpha_W=\alpha_B$, that is if both players are playing
+equally passively, then $\Pi_W=E_W$ and $\Pi_B=E_B$; their payoffs
+remain the same as in the standard Elo system, as they are \textit{both}
+playing equally suboptimally.
+
+Importantly, scaling $Q_W,Q_B$ by $\alpha_W,\alpha_B$ is equivalent to changing
+each player's effective rating by $R_W + 400\log_{10}(\alpha_W),R_B + 400\log_{10}(\alpha_B)$.
+This fact is used in the code when estimating $\alpha_A, \alpha_B$.
+
+This difference in rating is not the only effect that $\alpha_A,\alpha_B$ have,
+since this modified Elo description deals only with suboptimal play and changing expected values;
+the issue of changing draw probability will be addressed in the next section.
+
+\subsection{Estimating outcome probabilities}
+To estimate $p=(\pw,\pd,\pl)$, we initially use a multinomial regression as follows.
+\begin{gather*}
+    \bm{x} = (R_W, R_B)\\
+    \pw = \frac{\exp(-\bm{\beta}_w \cdot \bm{x})}
+    {1+\exp(-\bm{\beta}_w \cdot \bm{x}) + \exp(-\bm{\beta}_d \cdot \bm{x})} \\
+    \pd = \frac{\exp(-\bm{\beta}_d \cdot \bm{x})}
+    {1+\exp(-\bm{\beta}_w \cdot \bm{x}) + \exp(-\bm{\beta}_d \cdot \bm{x})} \\
+    \pl = 1 - \pw - \pd
+\end{gather*}
+To account for $\alpha_A, \alpha_B$ changing draw probabilities, we make the following
+modifications to $\pw,\pd,\pl$,
+beginning with the rating adjustments as described in the previous section.
+\begin{gather*}
+    R_{W_{\text{adj}}} = R_W + 400\log_{10}(\alpha_W), R_{B_{\text{adj}}} = R_B + 400\log_{10}(\alpha_B) \\
+    \bm{x}'=(\Rw, \Rb)\\
+    \sigma(x) = \frac{1}{1+\exp(-x)} \\
+    \gamma(\alpha_W, \alpha_B) = 1 - \frac{\sigma(\alpha_W)\alpha_W + \sigma(\alpha_B)\alpha_B}
+    {\sigma(\alpha_W)+\sigma(\alpha_B)} \\
+    p_{\text{win}_{\text{adj}}} = \frac{\exp(-\bm{\beta}_w \cdot \bm{x}')}
+    {1+\exp(-\bm{\beta}_w \cdot \bm{x}') + \exp(-\bm{\beta}_d \cdot \bm{x}')} \\
+    p_{\text{draw}_{\text{adj}}} = \frac{\exp(-\bm{\beta}_d \cdot \bm{x})}
+    {1+\exp(-\bm{\beta}_w \cdot \bm{x}') + \exp(-\bm{\beta}_d \cdot \bm{x}')}
+    + \gamma(\alpha_W,\alpha_B)\left( \pw - p_{\text{win}_{\text{adj}}} \right)\\
+    p_{\text{lose}_{\text{adj}}} = 1- p_{\text{win}_{\text{adj}}} - p_{\text{draw}_{\text{adj}}}
+\end{gather*}
+The function $\gamma$ is a sigmoid-weighted average of the two $\alpha$ parameters.
+It represents the fraction of draw probability that is gained by sacrificing win percentage.
+In other words, it represents the decrease in variance that comes at the cost of decreasing
+expected value.
+
+If both players choose the same $\alpha$, then the weighted average
+will come out to just $\alpha$, otherwise the average biases toward
+the player playing more optimally (that is, the player with the greater $\alpha$).
+That is, it is more punishing to the player playing more suboptimally passively; the
+player playing worse does not receive a proportional increase in draw percentage.
+
+\subsection{Defining $\alpha_W,\alpha_B$}
+For modelling player behavior, the following feature variables were used, inspired by David Smerdon's
 \href{https://www.davidsmerdon.com/?p=2168}{Fighting Chess Index}.
-Variables computed for each player were
+Variables computed for each player in the dataset were
 \begin{itemize}
@@
 \end{itemize}
-See Section 12.2.4 in Bishop's \textit{Pattern Recognition and Machine Learning} for a detailed description of this factor analysis model.
+Smerdon uses these features as inputs to PCA to compute an index of player aggression.
+Here, we use these features to estimate $\alpha_W, \alpha_B$.
 
-Smerdon uses PCA to form his Fighting Chess Index, but this method has its flaws. As described in Bishop,
-PCA can be formulated as a maximum likelihood solution of a latent variable model.
-But PCA does not treat the observed variables as conditionally independent given the latent variables.
-So if we use principal components to measure ``aggression'' or ``fighting spirit,'' we are not treating
-average rating (a measure of player strength) and irrational draw behavior as independent.
-This is problem since they are correlated.
-More serious (stronger) players have more on the line, and so will generally play for draws more often.
-But as we want to measure ``fighting spirit'' for individual player independent of their strength, we do not want this dependence.
+We add two vectors of weights $\bm{w}_W,\bm{w}_B$ to the model to represent
+linear combinations of the above features, $\bm{f}_W,\bm{f}_B$ for white and black respectively, with
+intercept terms $\delta_W,\delta_B$.
+Importantly, all features above were $z$-scored.
+\begin{gather*}
+    \alpha_W=\delta_W + \bm{w}_W\cdot \bm{f}_W \\
+    \alpha_B=\delta_B + \bm{w}_B\cdot \bm{f}_B
+\end{gather*}
 
-In Smerdon's defense, PCA's components do lend themselves to the construction of his index (whereas the factor model used here does not), but this does not change the problem of the underlying assumption.
+\subsection{Learning parameters}
+There are no hyperparameters in the model. All parameters
+$\bm{\beta}_d,\bm{\beta}_w, \delta_W, \delta_B,\bm{w}_W,\bm{w}_B$ are learned through maximum likelihood estimation
+according to the computed multinomial distribution
+$p=(p_{\text{win}_{\text{adj}}},p_{\text{draw}_{\text{adj}}},p_{\text{lose}_{\text{adj}}})$.
 
-The factor model used here does treat observations as conditionally independent of the latent variables.
-This is tremendously useful, since we want to retain measures of player strength while independently encoding
-draw behavior, especially as inputs to the following model.
-
-\subsection{Game model}
-(Again I will provide more detail wrt inference, etc in final, in a rush to the deadline as a result of repeated disasters)
-
-For each game, we modelled $p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$
-using the following generative model, where $l_w, l_b$ are the latent variables
- computed from the player model for the corresponding players.
-\begin{align*}
-    p = A l_w + B l_b + \epsilon \\
-    \epsilon \sim \mathcal{N}(\mathbf{0}, \sigma\mathbf{I}) \\
-    \sigma\sim \Gamma(\alpha=1,\beta=2)
-\end{align*}
+For more information see the GitHub repo.
 
 \section{Results}
+\subsection{Data}
+The dataset used was \href{http://caissabase.co.uk/}{Caissabase}, a free database of 5.61 million professional chess games.
+Simul games, Chess960 (a chess variant) games, and consultation (team) games were removed.
+Games played by players with fewer than 10 games were also removed.
+Games played by chess engines were also removed, as were games with invalid player names.
 
+Player names were cleaned and standardized using Levenshtein distance fuzzy string matching,
+see \href{https://github.com/seatgeek/thefuzz}{this Python package} for more details.
+
+This left 4,695,527 games in the dataset.
+
+\subsection{Model selection}
+To test the statistical significance of all the additional parameters, a parsimonious model
+was formulated that excludes $\alpha_W,\alpha_B$ and all additional parameters, instead only
+taking $R_W,R_W$.
+
+A likelihood ratio was computed followed by a chi-squared test, yielding a $p$-value of $<0.000001$.
+This is not unexpected, since those parameters describe draw behavior, whereas ratings and the Elo system more generally, as discussed
+below, do little to describe draw probabilities.
+
+\subsection{Distributions of $\alpha_W, \alpha_B$}
+
+Below are kernel density estimates of the densities of $\alpha_W,\alpha_B$.
+
+\includegraphics[width=\textwidth]{./distributions.png}
+
+Interestingly, the density at $\alpha_W=\alpha_B=1$ is almost $0$.
+This might suggest that no players are playing perfectly optimally, but more likely than not,
+this indicates one of the flaws of the Elo rating system.
+
+It is clearly useful (and very effective) for ranking players, but does poorly when applied
+to draw prediction. This is not relevant for its main purpose, but poses a challenge here.
+Because the Elo expected value $$E_W=\pw + \frac{1}{2}\pd$$
+neither defines $\pw$ nor $\pd$, instead assigning a ``payoff'' of $\frac{1}{2}$ to drawing,
+it becomes difficult to predict draw behavior using it.
+This is especially visible at high levels of play where draws are quite common;
+35\% of games in the cleaned dataset were draws.
+This flaw was also present in the
+modified Elo expected value, which was more to
+derive relative playing strength in the form of adjusted ratings, rather than directly compute draw probabilities.
+
+So rather than looking at the individual distributions of $\alpha_W,\alpha_B$, it is more informative to examine
+the distribution of their difference.
+
+\includegraphics[width=\textwidth]{./diff.png}
+
+This is almost zero-mean, with the average difference being $0.0015$ and the median difference being $0.0055$.
+It should be expected that white plays less passively, since white wins 37\% of the time as compared
+to black's 28\%
+(in this dataset).
+Interestingly the tails of the $\alpha_B$ distribution are quite long in comparison to the short tails of the $\alpha_W$ distribution.
+This could again be due to black's native disadvantage encouraging more deviation and more strategies chosen
+to either really attempt to win (higher $\alpha_B$) or attempt to draw at at all costs (lower $\alpha_W$).
+
+\includegraphics[width=\textwidth]{./linear.png}
+
+There is also a correlation between rating and $\alpha$, with an $R$-value of 0.5047, indicating that at higher ratings
+there is less suboptimal play.
+This is not totally unsurprising, since at the highest echelons of play this suboptimal playing behavior is much more easily
+observed, socially discouraged, and more heavily punished by other players (due to the higher levels of play).
+At slightly lower levels of play, from national masters to lower-rated international masters, though, we observe lower
+values of $\alpha$.
+
+\section{Conclusion}
+These results relate back to the article/real-world situation in a couple of ways.
+First, it shows that variance-lowering behavior at the cost of expected value does occur in chess.
+That is, playing suboptimally for draws at higher levels of play occurs often.
+It is difficult to identify, however, if this is as a result of random noise or variance in playing strength
+on behalf of players, or if each choice of $\alpha_W, \alpha_B$ for each game is intentional.
+It is also clear that this behavior, on average, decreases as the level of play increases, as predicted.
+
+That being said, the majority of situations involve players playing similarly, that is choosing
+$\alpha_W\approx \alpha_B$, so although some suboptimal hyperpassive play occurs, it does not occur
+sufficiently to pose a threat to the integrity of the game.
+
+The article mentions measures designed to increase errors and decrease the likelihood of draws.
+These would likely be successful at the highest echelons of play, where games are sufficiently
+scrutinized and play is so precise that suboptimal play essentially does not exist.
+However, at slightly lower levels of play (still professional, but just among weaker players), the problem is not absolute precision.
+Rather, (assuming the results support this), the problem is this variance-reducing behavior.
+So from a design perspective, the solution would be to penalize this behavior.
+Making draws worth $\frac{1}{3}$ of a point could discourage taking draws in tournaments, for instance.
+Removing agreed draws (where players simply agree to a draw rather than playing it out) as a whole could also make this behavior more difficult.
+
+% TODO BIB
+Caissabase, Elo, Smerdon, The Fuzz, original article
 \end{document}

### 2022-03-08 notebooks/logistic.R

@@ -1,0 +1,110 @@
+sigma <- function(x, k=1) 1 / (1 + exp(-k * x))
+
+gamma <- function(alpha_W, alpha_B) {
+    return(1 - (sigma(alpha_W) * alpha_W + sigma(alpha_B) * alpha_B) /
+           (sigma(alpha_W) + sigma(alpha_B)))
+}
+
+pred <- function(model, R_W, R_B) {
+    coefs <- as.vector(model$coefficients)
+    lin_term <- cbind(rep(1, length(R_W)), R_W, R_B) %*% coefs
+    return(exp(-lin_term))
+}
+
+log_likelihood <- function(par, R_W, R_B) {
+    intercept_W <- par[1]
+    intercept_B <- par[2]
+    weights_W <- par[3:(n_features + 2)]
+    weights_B <- par[(n_features + 3) : (2 + n_features * 2)]
+
+    alpha_W <<- sigma(intercept_W + white_features %*% weights_W)
+    alpha_B <<- sigma(intercept_B + black_features %*% weights_B)
+
+    R_W_adj <- R_W + 400 * log10(alpha_W)
+    R_B_adj <- R_B + 400 * log10(alpha_B)
+
+    t_w <- pred(w_model, R_W_adj, R_B_adj)
+    t_w0 <- pred(w_model, R_W, R_B)
+    t_d <- pred(d_model, R_W_adj, R_B_adj)
+    p_w0 <<- t_w0 / (1 + t_w0 + t_d)
+    p_w <<- t_w / (1 + t_w + t_d)
+    p_d <<- t_d / (1 + t_w + t_d) + gamma(alpha_W, alpha_B) * (p_w - p_w0)
+    p_l  <<- pmax(1 - p_d - p_w, rep(0, length(R_W))) # catch numerical errors
+
+    wins <<- data$Result == "1-0"
+    losses <<- data$Result == "0-1"
+    draws <<- data$Result == "1/2-1/2"
+
+    loglik <- sum(log(c(p_w[wins], p_d[draws], p_l[losses])))
+    return(loglik)
+}
+
+parsim <- function(R_W, R_B) {
+    t_w <- pred(w_model, R_W, R_B)
+    t_d <- pred(d_model, R_W, R_B)
+    p_w <<- t_w / (1 + t_w + t_d)
+    p_d <<- t_d / (1 + t_w + t_d)
[…]

### 2022-03-08 paper/paper.tex
final? changes

@@
 
-\section*{Disclaimer}
-I spent most of my time trying to get the models working. It was a complete nightmare. I thought it was close to working for many hours. Last 1\% took 99\% of the time.
-Getting the factor analysis model to work (replacing Smerdon's PCA) took me about 10 hours alone.
-Maybe I should have settled for PCA for this draft.
-Perhaps I should not have tried to relearn Pyro, a probabilistic programming language written in Python.
-That being said, if I had to build this model in R I likely would have ended up using RStan, which is a binding to another PPL, so the model wouldn't have really been in R either.
-The data cleaning and manipulation would also have taken me 10x longer since I don't really know R.
-Hindsight is 20/20. I think it's on its way, though.
+\section*{Description of changes}
+I changed all the models to be in principally game-theoretic terms.
+The models were rewritten, much of the code has changed, but the basic ideas
+and underlying assumptions have remained.
+The Elo rating system is still used, but I also modified it to allow it to represent
+suboptimal play and inform the predictive logistic regressions that are described later.
 
-The code (and this writeup) is in a rough state as a result -- by the final draft it will be pretty and commented and clear (despite being in Python).
-Similarly, since the model was giving me issues I didn't have compute time to train the model on the whole dataset (which is 4.6M games after cleaning).
-Again this will not be the case by the final draft. Everything works though; I have a complete pipeline
-that just needs some clean up, some hyperparameters to be adjusted/played with, and to be run to completion.
-
-Also, there are a couple of things I would like to add (or at least investigate) by the final draft:
-\begin{itemize}
-    \item Dataset of online games (ostensibly less incentive to draw here, more ``naturalistic'' or aggressive gameplay)
-    \item Using the factor model to compute an index, comparing to Smerdon's results
-\end{itemize}
+The only piece of feedback I was unable to incorporate was running simulations to generate
+data (though I could have done this after training the regressions, it would not have been useful).
+This is principally because of the way the Elo system works, and its description of an expected
+value rather than as probabilities of winning, drawing, or losing.
+So without making baseless assumptions (that would have resulted in useless simulations), generating
+simulated data would not have been useful.
+There was also plenty of real data to work with as well.
 
@@
 \section{Conclusion}
-These results relate back to the article/real-world situation in a couple of ways.
-First, it shows that variance-lowering behavior at the cost of expected value does occur in chess.
+These results relate back to \href{https://fivethirtyeight.com/features/some-humble-suggestions-to-save-chess-from-itself/}{Roeder's article}/the real-world situation in a couple of ways.
+
+First, they shows that variance-lowering behavior at the cost of expected value does occur in chess.
 That is, playing suboptimally for draws at higher levels of play occurs often.
@@
 
-The article mentions measures designed to increase errors and decrease the likelihood of draws.
-These would likely be successful at the highest echelons of play, where games are sufficiently
-scrutinized and play is so precise that suboptimal play essentially does not exist.
+Roeder mentions a few measures designed to decrease drawing behavior, but these again are for
+the highest echelons (super GM levels) of play.
+These measures increase errors and decrease the likelihood of draws, whether by shortening time
+controls, forcing tiebreakers to be played before matches (to force playing for wins), and so on.
+These would likely be successful at the highest levels of play, where games are sufficiently
+scrutinized and play is so precise that suboptimal play essentially cannot exist.
+
 However, at slightly lower levels of play (still professional, but just among weaker players), the problem is not absolute precision.
@@
 
-% TODO BIB
-Caissabase, Elo, Smerdon, The Fuzz, original article
+\begin{thebibliography}{9}
+
+\bibitem{caissabase}
+\href{http://caissabase.co.uk/}{Caissabase}. Unknown author.
+
+\bibitem{elo}
+Elo, Arpad (1986). \textit{The Rating of Chessplayers, Past and Present}.
+
+\bibitem{polgar}
+Polgar, Susan (2011). \href{https://chessdailynews.com/the-700-lbs-gorilla-issue-to-draw-or-not-to-draw/}{The 700 lbs gorilla issue: To draw or not to draw.} \textit{Chess Daily News}.
+
+\bibitem{538}
+Roeder, Oliver (2021). \href{https://fivethirtyeight.com/features/some-humble-suggestions-to-save-chess-from-itself/}{Some Humble Suggestions to Save Chess From Itself}. \textit{538}.
+
+\bibitem{thefuzz}
+SeatGeek (2022). \href{https://github.com/seatgeek/thefuzz}{TheFuzz}, Python package.
+
+\bibitem{fci}
+Smerdon, David (2021). \href{https://www.davidsmerdon.com/?p=2168}{\textit{The Fighting Chess Index}}.
+
+\end{thebibliography}
 \end{document}
